%==============================================================
%  Abstract
%==============================================================
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

This thesis studies item-to-item recommendation for IndiaMART, a large
business-to-business marketplace whose catalogue spans tens of millions of
products. The setting is defined by scale, extreme sparsity of the
product-to-product interaction data, purely implicit feedback, and a persistent
cold-start problem, against an incumbent production recommender --- a simple
frequency-based model with backfilling --- that is a deliberately strong
baseline. Three complementary lines of work are developed. First, product
navigation is modelled as a $K$-step Markov chain whose higher-order transitions
are computed through a low-rank matrix factorization; this saves several orders
of magnitude in memory, but an empirical evaluation shows that the factorization
does not preserve recommendation quality relative to the raw transition matrix or
the production baseline, and a structural limitation of weighted non-negative
matrix factorization is identified. Second, position bias --- the tendency of
users to click higher-ranked items regardless of relevance --- is modelled under
the examination hypothesis and estimated by three independent routes that agree
qualitatively; correcting for it by re-weighting transitions in the existing
recommendation engine improves click-through rate by roughly $3.5$ to $4$
percentage points consistently on held-out weeks. Third, a hierarchical
cosine-softmax embedding model is developed that cold-starts product embeddings
from text metadata and refines them from interactions, combining content and
collaborative signals in a single vector space searched by approximate nearest
neighbours at full catalogue scale; the interaction training adds clear signal
over a content-only baseline, though the model does not overtake the
frequency-based incumbent. A theme runs through all three: offline optimisation
metrics --- reconstruction error and negative log-likelihood --- are unreliable
proxies for recommendation quality, and in two cases the training objective and
the click-through rate move in opposite directions. The thesis reports these
findings together with the estimation methods, scaling techniques, and
evaluations that produced them.
