%==============================================================
%  Conclusion
%==============================================================
\chapter{Conclusion}
\label{ch:conclusion}

This thesis approached item-to-item recommendation for IndiaMART from three
directions --- a factorized $K$-step Markov chain, a position-bias correction,
and a learned embedding space --- each measured against the same demanding
yardstick, the production frequency-based model with backfilling. This closing
chapter draws the findings together and points to where the work can be taken
next.

\section{Summary of Findings}
\label{sec:concl-summary}

\paragraph{Matrix factorization for the $K$-step Markov recommender.}
Modelling navigation as a Markov chain and reading recommendations from the
powers of its transition matrix is attractive, but the powers are infeasible to
form at catalogue scale. Expressing them through a low-rank factorization reduces
the memory required for higher-order transitions from petabytes to tens of
gigabytes. The saving is real, yet it does not buy recommendation quality:
across subcategories the factorized recommendations trailed both the raw
transition matrix and the production baseline, the gap widening as the transition
order grew and reconstruction error accumulated, and a self-transition term did
not close it. Weighted NMF, which fits only the observed entries, was found to
have a structural weakness --- its unconstrained, unobserved entries inflate and
come to dominate the Markov chain --- that a near-zero reconstruction error on
observed entries conceals rather than resolves.

\paragraph{Position-bias estimation and correction.}
Raw click logs conflate relevance with presentation, and the effect is plainly
visible in the IndiaMART data, where clicks fall monotonically with rank. Under
the examination hypothesis, position bias was estimated by three independent
routes --- perfect-exposure aggregation, anchoring to external search sources,
and a controlled A/B experiment --- which agreed qualitatively and confirmed the
strong advantage of the top positions. Folding the correction into the existing
$K$-step engine as per-transition-type weights, with no change to the engine
itself, raised click-through rate by roughly $3.5$ to $4$ percentage points
consistently on held-out weeks. Of the three lines of work, this correction gave
the clearest and most direct improvement.

\paragraph{Embedding-based recommendation.}
A hierarchical cosine-softmax model learned a dense embedding per product,
cold-started from text metadata and refined from interactions, and served it
through approximate nearest-neighbour search at full catalogue scale. The
interaction training added clear signal over a content-only baseline, but the
learned embeddings did not overtake the frequency-based incumbent with
backfilling. The most striking result was diagnostic rather than absolute: the
training objective proved an unreliable guide to recommendation quality. Random
initialisation reached a lower loss than text initialisation yet recommended
worse, and on the industry subset the flat objective beat the hierarchical one on
loss while losing to it, and degrading, on click-through rate.

\paragraph{A cross-cutting lesson.}
The same caution recurred in two of the three lines of work: driving down an
offline optimisation metric --- reconstruction error or negative log-likelihood
--- did not reliably improve, and sometimes worsened, the recommendations
actually produced. The simple frequency-based model with backfilling remained a
strong baseline throughout. Taken together, these results argue for evaluating
recommendation methods on recommendation quality directly, and for treating a
lower training loss as, at best, a hypothesis to be checked rather than a result
in itself.

\section{Limitations and Future Work}
\label{sec:concl-future}

Several directions follow naturally from this work.

\begin{itemize}
  \item \textbf{Controlling the unobserved entries in weighted NMF.} The
        limitation identified in Chapter~\ref{ch:markov-factorization} points to
        factorizations that regularise or otherwise constrain the reconstruction
        away from the observed support, so that the reconstructed matrix remains a
        well-behaved transition operator everywhere rather than only where data
        was seen.
  \item \textbf{Refreshing the assumed recommendations for bias evaluation.} The
        bootstrapping diagnostic of Chapter~\ref{ch:position-bias} showed the
        assumed recommendation lists drifting out of date as they saturate, with
        mismatches rising over time. Periodically refreshing them, and evaluating
        the effect of that refresh, is a concrete next step.
  \item \textbf{Closing the gap to the frequency baseline for embeddings.} The
        embedding model beats the content-only baseline but not the frequency
        incumbent. Combining the learned embeddings with the frequency signal, and
        applying the position-bias correction to the interaction weights that
        train them, are promising ways to combine the strengths of the three lines
        of work rather than treating them separately.
  \item \textbf{Broader experimental coverage.} The factorization experiments
        reported here are at subcategory scale; extending them, and the embedding
        evaluation, across more of the catalogue --- and to industry-level
        aggregations --- would sharpen the picture of where each method helps.
\end{itemize}

The unifying message of the thesis is that, in a marketplace of this scale and
sparsity, the value of a method is decided at the point of recommendation, not at
the point of optimisation. The position-bias correction succeeded because it
improved the signal the recommender actually consumes; the factorization and
embedding approaches, for all their formal appeal, were held back by the gap
between what they optimised and what they were meant to deliver. Narrowing that
gap is the thread that ties the future directions above together.
